---
title: "Toad"
author: "Andrew Palacios, Priscilla Ta, Anissa Stull"
format: 
  html:
    code-fold: true 
    toc: true
    number-sections: false
    embed-resources: true
editor: visual
execute:
  echo: true
  message: false
  warning: false
---
::: panel-tabset
# Load Libraries and Clean Data

```{r}
library(tidyverse)
library(here)
library(janitor)
library(adehabitatHR)
library(lubridate)
library(dplyr)
library(tibble)
library(sp)
library(sf)
library(patchwork)

toad <- read_csv(here("data", "arroyotoad_presence.csv")) %>%
  clean_names()

# Rename column
toad <- toad %>%
  rename(date = observed_on)

# Convert date column to Date format (if not already)
toad$date <- as.Date(toad$date)

# Extract year from date
toad <- toad %>%
  mutate(year = year(date))

# select columns from dataset
toad_selected <- toad %>%
  dplyr::select(year, latitude, longitude, place_county_name, place_state_name, place_country_name, scientific_name)

# View the first few rows of the new dataset
head(toad_selected)

#export toad_selected as csv
#write_csv(toad_selected, here("data", "toad_selected.csv"))

```
# Spatial Data Processing in ArcGIS Pro
Isolating Cleveland National Forest Boundary
- Extracted Cleveland National Forest Boundary from USFS Administrative Forests Boundaries shapefile to create "clevelandNF.shp"

Isolating Relevant Fire Perimeters
- Clipped the CALFIRE Historic Fire Perimeters (1991-2023) shapefile to the Cleveland National Forest boundary
- Since we are only interested in the fires that occurred during the years we have toad presence data, we filtered the clipped fire perimeters to only include fires from 2004 to 2023 - creating a new layer called "fires_cleveland_04to23.shp"

Isolating Arroyo Toad Presences within Cleveland National Forest
- Intersect toad_selected.csv with "clevelandNF.shp" to create a new shapefile that only contains toad occurences within Cleveland National Forest ("cleveland_toad.csv")


# Toad Home Range Analysis
```{r}
# Load the data

cleveland_toad <- read_csv(here("data", "cleveland_toad.csv")) %>%
  clean_names()

# Convert dataset to SpatialPointsDataFrame
coordinates(cleveland_toad) <- ~longitude + latitude
proj4string(cleveland_toad) <- CRS("+proj=longlat +datum=WGS84")

# Kernel Density Estimation (KDE) Home Range
kde <- kernelUD(cleveland_toad, h="href")  
hr95 <- getverticeshr(kde, 95)  # 95% home range
hr50 <- getverticeshr(kde, 50)  # 50% core area

# Minimum Convex Polygon (MCP) Home Range
mcp95 <- mcp(cleveland_toad, percent = 95)

# Plot results
par(mfrow=c(1,2))
plot(hr95, col="lightblue", main="KDE Home Range (95%) & Core Area (50%)")
plot(hr50, col="red", add=TRUE)
points(cleveland_toad, pch=20, col="black")

plot(mcp95, col="lightgreen", main="Minimum Convex Polygon (95%)")
points(cleveland_toad, pch=20, col="black")

# Save as shapefiles
#st_write(st_as_sf(hr95), "homerange_95.shp")
#st_write(st_as_sf(hr50), "homerange_50.shp")
#st_write(st_as_sf(mcp95), "homerange_mcp95.shp")

```

# Mapping
```{r}
# Load shapefiles using here()
clevelandNF <- st_read(here("data", "shapefiles", "clevelandNF.shp"))  # Cleveland National Forest
burn_scars <- st_read(here("data", "shapefiles", "fires_cleveland_04to23.shp"))  # Fire burn scars

# Convert to sf 
cleveland_toad_sf <- st_as_sf(cleveland_toad)
hr95 <- st_as_sf(hr95)
hr50 <- st_as_sf(hr50)
mcp95 <- st_as_sf(mcp95)
clevelandNF <- st_as_sf(clevelandNF)
burn_scars <- st_as_sf(burn_scars)

# 1 KDE Home Range over Cleveland National Forest
p1 <- ggplot() +
  geom_sf(data = clevelandNF, fill = "lightgreen", color = "darkgreen", alpha = 0.5) +
  geom_sf(data = hr95, aes(geometry = geometry, fill = "95% KDE"), color = NA, alpha = 0.3) +
  geom_sf(data = hr50, aes(geometry = geometry, fill = "50% KDE"), color = NA, alpha = 0.5) +
  geom_sf(data = cleveland_toad_sf, aes(geometry = geometry, color = "Toad Observations"), size = 1, alpha = 0.7) +
  scale_fill_manual(name = "Home Range", values = c("95% KDE" = "blue", "50% KDE" = "purple")) +
  scale_color_manual(name = "Observations", values = c("Toad Observations" = "black")) +
  labs(
    title = "Toad Home Range in Cleveland National Forest",
    x = "Longitude", 
    y = "Latitude",
    caption = "Figure 1: Estimated home range of Cleveland National Forest toads using KDE. 
               95% KDE (blue) represents the broader home range, while 50% KDE (purple) 
               indicates the core area. Black points represent individual observations."
  ) +
  theme_minimal()

# 2 Burn Scars over Cleveland National Forest
p2 <- ggplot() +
  geom_sf(data = clevelandNF, fill = "lightgreen", color = "darkgreen", alpha = 0.5) +
  geom_sf(data = burn_scars, aes(geometry = geometry, fill = "Burn Scars"), color = "black", alpha = 0.5) +
  scale_fill_manual(name = "Fire History", values = c("Burn Scars" = "red")) +
  labs(
    title = "Burn Scars in Cleveland National Forest",
    x = "Longitude", 
    y = "Latitude",
    caption = "Figure 2: Historical burn scars (red) in Cleveland National Forest. 
               Burned areas may influence toad habitat and movement patterns."
  ) +
  theme_minimal()

# Arrange the two plots side by side
p1 | p2
```

# Quantify 
```{r}

# Ensure the toad occurrences are in the same CRS as burn scars
cleveland_toad_sf <- st_transform(cleveland_toad_sf, st_crs(burn_scars))

# Perform spatial intersection to find toad occurrences within burn scars
toad_in_burn <- st_intersects(cleveland_toad_sf, burn_scars, sparse = FALSE)

# Count occurrences inside burn scars
num_toads_in_burn <- sum(rowSums(toad_in_burn) > 0)

# Total number of toad occurrences
total_toads <- nrow(cleveland_toad_sf)

# Calculate percentage of toad occurrences in burn scars
percent_in_burn <- (num_toads_in_burn / total_toads) * 100

# Print the result
print(paste("Percentage of toad occurrences in burn scars:", round(percent_in_burn, 2), "%"))



```



:::
